---
title: "Conversion of metabolomics data and metadata to standardized format"
author: 
  - Anna Niehues^[Radboud University Medical Center, Anna.Niehues@radboudumc.nl]
  - Purva Kulkarni^[Radboud University Medical Center, Purva.Kulkarni@radboudumc.nl]
output: bookdown::html_document2
bibliography: bibliography.bib
---

## Input data

Read input data from supplementary table S1 of [@Su2020].

```{r}
require(readxl)
table_s1_path <- here::here("../../data/Su_2020_original", "1-s2.0-S0092867420314446-mmc1.xlsx")
excel_sheets(table_s1_path)
table_s1_mtbl <- read_excel(
  table_s1_path, sheet = "S1.4 Plasma Metabolomic Data")
head(table_s1_mtbl)
```

### Extract metabolites names as given in input data

```{r}
metabolites_original <- colnames(table_s1_mtbl)[-c(1,2)]
head(metabolites_original)
```


## biodbChebi

Search ChEBI database using metabolites names

https://www.bioconductor.org/packages/release/bioc/html/biodbChebi.html

https://www.bioconductor.org/packages/release/bioc/vignettes/biodbChebi/inst/doc/biodbChebi.html#8_Search_by_name_and_mass

```{r}
library(biodbChebi)
```

```{r}
# create instance of biodb class BiodbMain
mybiodb <- biodb::newInst()
# create connector to ChEBI database
chebi <- mybiodb$getFactory()$createConn("chebi")
chebi$checkDb()
```

The following step takes several minutes

```{r}
library("data.table")
metabolites_chebi_dfs <- lapply(metabolites_original, function(m) {
  #print(m)
  # search name in ChEBI
  chebi_ids <- chebi$searchCompound(name = m, max.results = 5)
  chebi_tmp_list <- list()
  num_tmp_list_entries <- 1
  chebi_entries <- lapply(chebi_ids, function(x) {
    # address possible error
    # try(chebi$getEntry(7944)) ERROR [15:01:28.295] Cannot set more that one value (C19H20FNO3.ClH, C19H21ClFNO3) into single value field "formula" for entry ChEBI 7944
    e <- NULL
    try(e <- chebi$getEntry(x), silent = TRUE) 
    if (is.null(e)) {
      # create entry with query name and ChEBI ID only
      chebi_tmp_list[[num_tmp_list_entries]] <<- data.frame(
        query = m, chebi.id = x)
      num_tmp_list_entries <<- num_tmp_list_entries + 1
    }
    e
  })
  chebi_entries <- chebi_entries[!sapply(chebi_entries,is.null)]
  if (length(chebi_entries) == 0) { NULL # no matches in ChEBI database
    } else {
    if (typeof(chebi_entries) == "environment") {
      # single entries are not returned as list - need to be converted
      chebi_entries <- list(chebi_entries)
    }
    # (first x) matching ChEBI entries as data frame
    chebi_df <- mybiodb$entriesToDataframe(chebi_entries)
      #fields = c("accession", "chebi.id", "inchi", "inchikey", "smiles",
      #           "kegg.compound.id", "molecular.mass", "monoisotopic.mass", 
      #           "charge", "formula", "n.stars", "name")
    chebi_df$query <- c(m) # add original metabolite name
    chebi_tmp_list[[num_tmp_list_entries]] <- chebi_df
    chebi_df <- rbindlist(chebi_tmp_list, fill = TRUE)
    chebi_df
  }
})
```


```{r}
require("data.table")
metabolites_df <- rbindlist(metabolites_chebi_dfs, fill = TRUE)
head(metabolites_df)
#write.csv(metabolites_df, here::here("data", "metabolites_ChEBI_entries.csv"))
```

```{r}
# terminate biodb instance
mybiodb$terminate()
```

```{r}
#metabolites_df <- read.csv(here::here("data", "metabolites_ChEBI_entries.csv"),
#                           row.names = 1, header= TRUE)
names(metabolites_df)
```

## BridgeDB

### Download BridgeDB database

See tutorial: https://www.bioconductor.org/packages/release/bioc/vignettes/BridgeDbR/inst/doc/tutorial.html#4_Metabolomics

See BridgeDB website: https://bridgedb.github.io/ 

Mapping database: https://doi.org/10.6084/m9.figshare.13550384.v1

```{r}
library(BridgeDbR)
#mtbl_db_url <- "https://figshare.com/ndownloader/files/26001794" # gives wrong md5, see below
mtbl_db_md5 <- "1b78be20f5fdc947dd15b9e8152879ce"
mtbl_db_path <- here::here("db", "metabolites_20210109.bridge")
# download metabolites database
#download.file(mtbl_db_url, mtbl_db_path) # this causes file to have wrong md5, downloaded manually - this worked
# check md5 checksum
hash <- tools::md5sum(mtbl_db_path)
stopifnot(hash == mtbl_db_md5)
# create a BridgeDB mapper (didn't work with file downloaded with download.file - which had wrong md5)
mapper <- loadDatabase(mtbl_db_path)
```

### Map metabolites

See https://bridgedb.github.io/pages/system-codes.html

| Code | Source |
|------|--------|
| Ca | CAS
| Ce | ChEBI
| Cs | ChemSpider
| Cl | ChEMBL compound
| Dr | DrugBank
| Ch | HMDB
| Gpl | Guide to Pharmacology Ligand ID (aka IUPHAR)
| Ik | InChIKey
| Ck | KEGG Compound
| Kd | KEGG Drug
| Kl | KEGG Glycan
| Lm | LIPID MAPS
| Lb | LipidBank
| Pgd | PharmGKB Drug
| Cpc | PubChem Compound
| Cps | PubChem Substance
| Sl | SwissLipids
| Td | TTD Drug
| Wd | Wikidata
| Wi | Wikipedia


Add HMDB identifiers

```{r}
metabolites_df$HMDB <- unlist(apply(metabolites_df, 1, function(row) {
  hmdb = c()
  # map ChEBI ID to HMDB
  df <- map(mapper, source = "Ce", identifier = row["chebi.id"], 
            target = "Ch")
  hmdb <- c(hmdb, df$mapping)
  # map KEGG Compound ID to HMDB 
  df <- map(mapper, source = "Ck", identifier = row["kegg.compound.id"], 
            target = "Ch")
  hmdb <- c(hmdb, df$mapping)
  if (length(hmdb) > 0) {
    # remove duplicates, merge to str
    hmdb <- paste0(unique(hmdb), collapse = ";") 
  } else {
    hmdb <- NA
  }
  hmdb
}))
```

```{r}
dir.create(here::here("../../data/Su_2020_intermediate/metabolomics"), 
                      showWarnings = FALSE)
write.csv(metabolites_df, here::here("../../data/Su_2020_intermediate/metabolomics", 
                                     "metabolites_ChEBI_entries.csv"))
```

## Filter results

```{r}
metabolites_df <- read.csv(here::here("../../data/Su_2020_intermediate/metabolomics", 
                                     "metabolites_ChEBI_entries.csv"),
                           row.names = 1)
```

Check if there is a complete match with one of the synonyms

```{r}
library(stringr)
# check if query matches one of the names in the chebi entry 'name'
# non-case-sensitive match
metabolites_df$complete.match.noncasesensitive <- unlist(
  apply(metabolites_df, 1, function(row) {
    complete_match = FALSE
    query <- str_to_lower(row["query"])
    synonyms <- sapply(strsplit(row["name"], split = ";"), str_to_lower)
    if (query %in% synonyms) {
      complete_match = TRUE
    }
    complete_match
}))

# check if query matches one of the names in the chebi entry 'name'
# case-sensitive match
metabolites_df$complete.match.casesensitive <- unlist(apply(
  metabolites_df, 1, function(row) {
    complete_match = FALSE
    query <- row["query"]
    synonyms <- unlist(strsplit(row["name"], split = ";"))
    if (query %in% synonyms) {
      complete_match = TRUE
    }
    complete_match
}))

# check if query matches the first name in the chebi entry 'name'
# case-sensitive match
metabolites_df$complete.match.firstentry <- unlist(apply(
  metabolites_df, 1, function(row) {
    complete_match = FALSE
    query <- row["query"]
    synonyms <- unlist(strsplit(row["name"], split = ";"))
    if (!is.na(synonyms[[1]])) {
      if (query == synonyms[[1]]) {
        complete_match = TRUE
      }
    }
    complete_match
}))

# remove whitespace and trailing strings in parentheses (e.g. abbreviations) from query
strip_query <- function(x) {
  # remove all whitespace
  partial <- str_replace_all(x, "[:space:]", "")
  # partial string - remove parentheses end content at end of string
  inparentheses <- str_extract(partial, "[ ]?\\([:print:]+\\)")
  if (!is.na(inparentheses)) {
    if (endsWith(partial, inparentheses)) {
      partial <- str_extract(partial, "[:print:]+(?=[ ]?\\([:print:]+\\))")
    }
  }
  partial
}

# score for string match between query metabolite name and ChEBI db synonyms ('name' in chebi entry)
# to rank entries, in case of multiple entries for one query
metabolites_df$match.score <- unlist(apply(metabolites_df, 1, function(row) {
  match_score <- 0
  query <- row["query"]
  query.lower <- str_to_lower(query)
  query.stripped <- strip_query(query.lower)
  synonyms <- sapply(strsplit(row["name"], split = ";"), str_to_lower)
  synonyms.lower <- sapply(synonyms, str_to_lower)
  synonyms.stripped <- sapply(synonyms.lower, strip_query)
  if (query.lower %in% synonyms.lower) {
    # query matches one of synonyms (ignoring upper/lower case)
    match_score <- match_score + 1
    if (query %in% synonyms) {
      # query matches one of synonyms (including match of upper/lower case)
      match_score <- match_score * 2
    }
    if (!is.na(synonyms[[1]])) {if (query.lower == synonyms.lower[[1]]) {
      # query matches first synonym
      match_score <- match_score * 2
    }}
  } else {
    if (query.stripped %in% synonyms.stripped) {
      # partial match 
      ## removing possible characters in parentheses at end of string
      ## these can be abbreviations or indicating charge)
      match_score <- match_score + 0.5
    }
  }
  match_score
}))

# score for annotations - prefer metabolites with KEGG compound ID and HMDB ID
metabolites_df$annotation.score <- unlist(apply(metabolites_df, 1, function(row) {
  annotation_score <- 0
  if (!is.na(row["kegg.compound.id"])) {
    annotation_score <- annotation_score + 1
  }
  if (!is.na(row["HMDB"])) {
    annotation_score <- annotation_score + 1
  }
  annotation_score <- annotation_score * as.integer(row["n.stars"])
  annotation_score
}))
```

In some cases molecule name synonyms of multiple entries have a complete match with the query name.

```{r}
head(metabolites_df)
names(metabolites_df)
View(metabolites_df[,c("name", "query", "match.score", "HMDB")])
```

Filter entries based on match score and annotation score

```{r}
require(data.table)
metabolites_df.filtered <- rbindlist(sapply(metabolites_original, function(m) {
  tmp_df <- metabolites_df[metabolites_df$query == m,]
  max_score <- max(tmp_df$match.score)
  if (max_score > 0) { # i.e. match
    tmp_df <- tmp_df[tmp_df$match.score == max_score,]
    num_best_matches <- dim(tmp_df)[1]
    if (num_best_matches == 1) {
      # single best hit based on match score - use single best hit
      best_match <- tmp_df[1,]
    } else if (num_best_matches > 1) {
      # multiple best hits with same score
      max_annotation <- max(tmp_df$annotation.score)
      tmp_df <- tmp_df[tmp_df$annotation.score == max_annotation,]
      num_best_annotation <- dim(tmp_df)[1]
      if (num_best_annotation == 1) {
        # use best annotated hit 
        best_match <- tmp_df[1,]
      } else {
        # skip query
        best_match <- NULL
        # use first hit
        # best_match <- tmp_df[1,]
        # print(best_match$name)
      }
    }
  } else {
    best_match <- NULL
  }
  best_match
}))
```

```{r}
write.csv(metabolites_df.filtered, 
          here::here("../../data/Su_2020_intermediate/metabolomics", 
                     "metabolites_ChEBI_entries_filtered.csv"))
```

## Create feature annotation file

Save filtered file with metabolites in rows and samples in columns

```{r}
names(table_s1_mtbl[1:3])
metabolites_filtered <- metabolites_df.filtered$query
length(metabolites_filtered)
table_s1_mtbl.filtered <- as.data.frame(table_s1_mtbl[, c("sample_id", metabolites_filtered)])
rownames(table_s1_mtbl.filtered) <- table_s1_mtbl.filtered$sample_id
table_s1_mtbl.filtered <- table_s1_mtbl.filtered[, 2:dim(table_s1_mtbl.filtered)[2]]
table_s1_mtbl.filtered_t <- as.data.frame(t(table_s1_mtbl.filtered))
table_s1_mtbl.filtered_t <- cbind(data.frame(`feature name` = rownames(table_s1_mtbl.filtered_t)),
            table_s1_mtbl.filtered_t)

dir.create(here::here("../../data/Su_2020_FAIR/metabolomics"), 
                      showWarnings = FALSE)
write.csv(table_s1_mtbl.filtered_t, 
          here::here("../../data/Su_2020_FAIR/metabolomics", 
                     "metabolomics_Su_2020_feature-data.csv"),
          row.names = FALSE)
```


### Create annotation file and save

```{r}
metabolites_annotations <- data.frame(
  `feature name` = metabolites_df.filtered$query,
  `database ID` = metabolites_df.filtered$chebi.id,
  `database` = c("https://www.ebi.ac.uk/chebi/"),
  `database version` = c("ChEBI Release 211")
)
```

```{r}
write.csv(metabolites_annotations, 
          here::here("../../data/Su_2020_FAIR/metabolomics", 
                     "metabolomics_Su_2020_feature-metadata.csv"),
          row.names = FALSE)
```


## References
